---
title: "Part 2"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

# Data Frames

We're going to work with data on penguins from https://github.com/allisonhorst/palmerpenguins.  It's been packaged into an R package.  We're going to install the remotes package that allows us to install R packages from github repositories.  

You'll likely be asked about updating packages during the installation.  You can skip the updates (just hit Return or enter the value for None).

```{r, installpackages, eval=FALSE}
install.packages("remotes")
remotes::install_github("allisonhorst/palmerpenguins")
```

```{r}
library(palmerpenguins)
```


## Manual Creation

How we could manually create a data frame (very rarely will do this)
```{r, manualdf}
x <- data.frame(month=month.name,
                index=1:12,
                days=c(31,28,31,30,31,30,31,31,30,31,30,31))
x
```

## Data Shape and Names

Use the penguin data
```{r, loaddata}
data("penguins")
```

What is `penguins`?
```{r}
class(penguins)
```

"tbl_df" is a tibble data frame.  These behave a little bit differently from normal data frames.  You'll see tibbles instead of data frames within the tidyverse set of packages (and those packages that work within that framework).

The biggest difference is that tibbles give you a tibble back when subsetting with [], while data frame sometimes give you a vector.

View the first few rows
```{r}
head(penguins)
```

Or user the viewer:
```{r, eval=FALSE}
View(penguins)
```

Clicking on the name of the data frame in the environment tab will also open the viewer.

Dimensions of the data frame
```{r}
dim(penguins)
```

What will the length of a data frame be?  # of columns
```{r}
length(penguins)
```

This is because it's technically a list of vectors (lists are a different type).  Don't use this.  Use `ncol()` instead:
```{r}
ncol(penguins)
```


```{r}
nrow(penguins)
```

What are the variable names?  The columns
```{r}
names(penguins)
```

Rows have names too
```{r}
rownames(penguins)
```

These were generated by default.  Generally don't use row names

What are the column types?
```{r}
str(penguins)
```

Quick summary of every columns
```{r}
summary(penguins)
```

Just look at the first few rows of data:
```{r}
head(penguins)
```

**EXERCISE** (Run the code above first that creates the `x` data frame): Using the `x` data frame we created above:
* How many rows and columns?
* Names of the variables?

```{r}

```


## Indexing

Select first row
```{r}
penguins[1,]
```


Select first two rows
```{r}
penguins[1:2,]
```


Select first column
```{r}
penguins[,1]
```

This gave us a tibble/data.frame back, but with `x`:
```{r}
x[,1]
```

we get a vector back.

**EXERCISE** (using `x` created above): how would we select rows 2 through 5, and columns 1 through 3 from `x`?
```{r}

```

Reference columns by name with `$` notation (no quotes on names)
```{r}
names(penguins)
```

```{r}
penguins$species
```
Note that the `$` notation got us a vector back.

```{r}
penguins$culmen_length_mm
```

Use names in `[]`: put them in quotes
```{r}
penguins[,"species"]
```

Since we used `[]`, we get a tibble/data.frame back this time
```{r}
penguins[,"culmen_length_mm"]
```

Multiple columns by name, need to use a vector of names:
```{r}
penguins[,c("island", "species")]
```

**EXERCISE**: Using the `x` data frame we created above:
* Select the days column using `$` notation
* Select the month and days columns by name using `[]`

```{r}

```



## Boolean Selection

If we have a boolean vector (`TRUE` and `FALSE` values) that is the same length as the number of rows or columns, we can use it to select from the data frame as we did with vectors.
```{r}
penguins[penguins$culmen_length_mm < 34,]
```

Note the rows of `NA`.

If I forget the `,` in the `[]`:
```{r, eval=FALSE}
penguins[penguins$culmen_length_mm < 34]
```

It tries to index the columns instead, and our vector is too long.

Multiple conditions
```{r}
penguins[penguins$culmen_length_mm < 34 & penguins$culmen_depth_mm < 16,]
```

```{r}
penguins[penguins$culmen_length_mm < 34 | penguins$culmen_length_mm > 58,]
```

**EXERCISE** Using `x` created above: select rows from `x` where the `month` has 31 days.  Remember: `==` for testing for equality.

```{r}

```



## Renaming Columns

```{r}
oldnames <- names(penguins)  # save so we can reset later
names(penguins)[1] <- "boo"  # change the name of the first column
names(penguins)
```

```{r}
names(penguins) <- c("a", "b", "c", "d", "e", "f", "g", "h")  # change all of the column names
head(penguins)
```

Put the old names back
```{r}
names(penguins) <- oldnames
head(penguins)
```


## Working with Variables

```{r}
max(penguins$culmen_length_mm)
```

```{r}
max(penguins$culmen_length_mm, na.rm=TRUE)
```

Now to get the observations (rows) for the penguins with that max value:
```{r}
penguins[penguins$culmen_length_mm == max(penguins$culmen_length_mm, na.rm=TRUE),]
```

Break it out into steps:
```{r}
max_culmen_length <- max(penguins$culmen_length_mm, na.rm=TRUE)
penguins[penguins$culmen_length_mm == max_culmen_length,]
```

**EXERCISE** Using `x` created above: select the rows from `x` where `days` is at its minimum value.
```{r}

```



## Making new variables

We can add new variables to the data frame by naming them with the `$` notation:
```{r}
penguins$culmen_length_cm <- penguins$culmen_length_mm / 100
penguins[, c("culmen_length_cm", "culmen_length_mm")]
```


**EXERCISE** Using `x` created above: make a new variable as part of `x`, called `weeks`, that is the number of days divided by 7

```{r}

```



## Missing values

Any missing values?  Show up in summary() output:
```{r}
summary(penguins)
```

Look at the rows where `body_mass_g` is missing:
```{r}
is.na(penguins$body_mass_g)
penguins[is.na(penguins$body_mass_g),]
```

Remove rows where `body_mass_g` is missing by selecting rows where the value is not missing:
```{r}
penguins[!is.na(penguins$body_mass_g),]
penguins <- penguins[!is.na(penguins$body_mass_g),]
```

Remove by index instead:
```{r}
penguins <- penguins[-1,]  # removes the first row of the data
```

**EXERCISE** Using `x` created above: remove the 2nd row of `x`:
```{r}

```


## Replacing Values

```{r}
table(penguins$species)
```

```{r}
penguins$species[penguins$species == "Gentoo"]  # which observations do we want to replace?
penguins$species[penguins$species == "Gentoo"] <- "gentoo"  # set the value
```


**EXERCISE** Using `x` created above: replace the `month` value "December" with the value "Dec" instead

```{r}

```



# Sort and Order

We're going to talk about sorting and ordering with a vector first.

```{r}
y <- c(30,60,20,30)
```

`sort()` puts the values in order (doesn't change the vector, returns a copy)
```{r}
sort(y)
```

```{r}
sort(y, decreasing=TRUE)
```

`order()` tells you what order to put the elements in to sort them:
```{r}
order(y)
```

You can use the output of `order()` to sort:
```{r}
y[order(y)]
```

`order()` also allows decreasing:
```{r}
y[order(y, decreasing=TRUE)]
```

Why do we care?  Because we need this if we want to sort a data frame.  `sort()` only take a vector -- one column of a data frame.  If we want to sort by one or more columns, we need to use `order()` to figure out what to put the rows in -- because we'd want entire rows re-ordered together.

We can sort a single column by itself
```{r}
sort(penguins$culmen_length_mm)
```

But if we want rows re-ordered by the value of culmen_length_mm, we need to get the order:
```{r}
order(penguins$culmen_length_mm)
```

and then use that to re-order the whole data frame
```{r}
penguins[order(penguins$culmen_length_mm), ]
```

`order()` will take multiple vectors (columns):
```{r}
penguins[order(penguins$island,
               penguins$culmen_length_mm), ]
```

**EXERCISE**: sort the `penguins` data frame by `flipper_length_mm` decreasing
```{r}

```


# Coming Up

Today you'll work more with data frames.  You'll also learn how to read a csv file into R into a data frame, and go over the parts of an R help page.

There are exercises, but I's stronly encourage you to read in one of your own data files and start working with it.  If you don't have a data file, there are some examples linked in the exercises that you could try.





