---
title: "Data Project"
date: "6/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Get the Data

For this project, you'll explore data from the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/).  It is a subset of the Illinois data from 2017 for Evanston Police.  You can learn about the data and the variables in the [project README file](https://github.com/stanford-policylab/opp/blob/master/data_readme.md).  Each observation is a stop by the police.

The subset is available in `data/ev_police.csv`.  The data may have been added after you downloaded the workshop materials from GitHub.  To read it directly from GitHub:

```{r}
police <- read.csv("https://github.com/nuitrcs/r-online-2020/raw/master/data/ev_police.csv",
                   stringsAsFactors = FALSE,
                   header=TRUE)
```

If you'd like to work with a larger set of data, you can download it from the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/).  Use the function `readRDS()` to open `.rds` files.  

You can use your skills to explore this data however you'd like.  Some suggestions are below.

You may want to reference [demographic information](https://datausa.io/profile/geo/evanston-il/#demographics) about Evanston for context for the data.

# Explore the Data

Get a list of the variable names in the data.

View the data.

How many rows are in the data?

Do any of the variables have missing data?

Which variables have useful data and which can you drop (because they have no variation or duplicate another column)?

Can you turn the `location` column into a more useful form?  Hint: the values are zip codes.  The main zip codes for Evanston are 60201 and 60202.  

# Possible Questions

Does the rate of citation, warning, or search vary across different types of violations?  by `subject_race`?

Are subjects of different races or sexes stopped at rates similar to the [demographics for Evanston](https://datausa.io/profile/geo/evanston-il/#demographics)?  

Is the age or make of the car related to the outcome of the stop? or the reason for the stop?

Are more citations given on certain days than others?  You may want to look at the lubridate package function `wday()` to help with this.


# Exploring Further

Interested in Evanston policing data beyond vehicle stops?  See https://data.cityofevanston.org/browse?category=Police.  




